\chapter{Project Management}
\label{ch:mgmt}

Applicable regulations, intellectual-property considerations, and the
dataset-licensing analysis are discussed in \Cref{ch:sustain}.

% ── Gantt colour palette (reused from the GEP deliverable) ─────────────────────
\definecolor{gpMgmt}{HTML}{1565C0}
\definecolor{gpSprint1}{HTML}{2E7D32}
\definecolor{gpSprint2}{HTML}{E65100}
\definecolor{gpSprint3}{HTML}{880E4F}
\definecolor{gpSprint4}{HTML}{6A1B9A}
\definecolor{gpSprint5}{HTML}{00695C}
\definecolor{gpMilestone}{HTML}{D32F2F}

% ─────────────────────────────────────────────────────────────────────────────
\section{Justification of the Chosen Architecture}
\label{sec:mgmt-justification}

\Cref{ch:soa} surveys the three principal families of OMR
systems in the literature (traditional modular pipelines,
CRNN--CTC sequence models, and transformer-based end-to-end
models), and \Cref{ch:design} develops the technical case for
selecting the two-stream CRNN--CTC system used in this thesis.
This section justifies that choice from the project-management
standpoint, by applying six criteria that combine the technical
target with the practical constraints of a Bachelor's thesis,
and by recording the schedule and budget consequences of the
decision.

\subsection{Selection criteria}

\textit{Expected accuracy on monophonic lines.}
Published accuracy on monophonic input must be competitive with the state of the art.

\textit{Computational feasibility on consumer hardware.}
Training must fit within the 12\,GB VRAM of the available NVIDIA RTX~3060 without cloud rental.

\textit{Data requirements.}
The architecture must be trainable on PrIMuS~\cite{PrIMuSDataset} ($\approx$88\,000 incipits), the only substantial labelled corpus for monophonic OMR.

\textit{Implementation complexity.}
A single-person project requires a well-documented architecture with mature open-source reference implementations.

\textit{Reproducibility within the project's timeline.}
Each sprint must produce a measurable, reproducible result; architectures requiring dozens of trial runs before convergence are impractical.

\textit{Extension potential for future work.}
The chosen architecture should support natural extensions to polyphony, full-page input, and chord-symbol integration.

\begin{table}[ht]
    \centering
    \caption{Qualitative comparison of architectural alternatives
    against the six project criteria.
    Symbols denote relative suitability: \textbf{++} excellent,
    \textbf{+} good, \textbf{--} poor, \textbf{-{}-} very poor.}
    \label{tab:alt-comparison}
    \small
    \setlength{\tabcolsep}{5pt}
    \renewcommand{\arraystretch}{1.35}
    \begin{tabularx}{\linewidth}{@{} >{\raggedright\arraybackslash}p{4.2cm} c c c @{}}
        \toprule
        \textbf{Criterion} & \textbf{Traditional pipeline} & \textbf{CRNN--CTC} & \textbf{Transformer} \\
        \midrule
        Monophonic line accuracy          & --  & +  & ++ \\ \tfgrowsep
        Computational feasibility         & ++  & +  & -{}- \\ \tfgrowsep
        Data requirements                 & ++  & +  & -{}- \\ \tfgrowsep
        Implementation complexity         & +   & +  & -{}- \\ \tfgrowsep
        Reproducibility in TFG timeline   & +   & ++ & -- \\ \tfgrowsep
        Extension potential (future work) & --  & +  & ++ \\
        \bottomrule
    \end{tabularx}
\end{table}

\Cref{tab:alt-comparison} shows that CRNN--CTC is the best fit
for this stage: it provides enough modelling capacity for monophonic
lead sheets while remaining trainable and debuggable within the
available timeframe and hardware budget.
The transformer alternatives are penalised principally by the data
and compute requirements of a from-scratch training run on a single
consumer GPU; a traditional modular pipeline is cheap to implement
but is penalised by error propagation across stages and by its
sensitivity to the jazz engraving style.
The latter is retained as the \textit{lower-bound}
baseline against which the CRNN--CTC results are compared in
\Cref{ch:results}.

\subsection{Schedule and budget consequences}

This choice shaped the project plan in several ways.
The training corpus could be built entirely from a single
public dataset (PrIMuS) re-rendered in the target engraving style,
which kept the data-acquisition cost within the thesis budget.
The output representation is a compact flat token
sequence (LMX, \Cref{sec:lmx-format}) consumed by both the
music and the chord branches, avoiding the verbose-XML output that
transformer-decoder architectures typically produce.
Above all, training time on a consumer GPU is short enough
(of the order of a day per full run) that several iterations per
sprint are realistic, which is what makes the iterative methodology
of \Cref{sec:mgmt-methodology} workable in practice.

The decision is also a risk-control choice: implementing the
traditional baseline early (WP1) guarantees measurable results
regardless of CRNN--CTC convergence, and the conditional
extension task T5.3 absorbs schedule overrun from the more
ambitious extensions (see \Cref{sec:mgmt-planning,sec:mgmt-risks}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Methodology}
\label{sec:mgmt-methodology}

The project follows an \textbf{iterative and incremental}
methodology with sprint-based checkpoints and supervisor reviews.
Three alternatives were weighed against this choice: a waterfall
plan offers no defined point at which per-token error analysis
can revise an earlier phase; Kanban removes the cadenced
checkpoints that biweekly supervisor reviews require; and full
Scrum imposes team ceremonies that presume a multi-person setup.
The iterative-incremental variant adopted here keeps Scrum's
sprint cadence without the team scaffolding, and preserves the
data-model-evaluation feedback loop that a waterfall plan would
have suppressed.
Each cycle produces measurable artefacts (datasets, training
logs, SER values, error analyses), creating an audit trail that
grounds every replanning decision in evidence and gives the
supervisor concrete material to react to at each biweekly
meeting.

Execution is organised as a closed loop between three activities:
\textit{data construction} (rendering, augmentation, vocabulary),
\textit{model training} (architecture changes, hyperparameter sweeps,
training-curve monitoring), and \textit{evaluation and error analysis}
(SER measurement, per-token breakdown, qualitative case studies).
An error analysis that surfaces a systematic failure mode triggers a
data-pipeline change, which triggers a new training run, whose results
feed back into the next error analysis.
The work-package ordering WP1 $\to$ WP2 $\to$ WP3 $\to$ WP4 $\to$ WP5
describes only the first traversal of this loop; sprint reviews
regularly send the project back to a prior WP for incremental
refinement, and the budget reserves time for that.

Four practices enforce methodological rigour across the project.
\begin{itemize}
    \item \textit{Traceability.} Every artefact (dataset version, checkpoint, SER value) is linked to a repository commit and a documented design decision, so that any number in \Cref{ch:results} can be traced to the data and code that produced it.
    \item \textit{Reproducibility.} Data generation, training, and evaluation are script-driven; the mechanism (embedded \code{Config}, deterministic seeds, checkpoint format) is described in \Cref{sec:impl-reproducibility}.
    \item \textit{Quantitative control.} SER is measured at every significant training iteration alongside per-token breakdown and per-sample distribution, so no decision rests on a single aggregate number (\Cref{sec:res-errors}).
    \item \textit{Corrective loop.} Each sprint ends with a supervisor review that adjusts scope, effort estimates, or risk responses via the budget-control mechanism of \Cref{sec:mgmt-control}.
\end{itemize}

Monitoring tools include Git and GitHub for version control,
per-experiment training logs and CSV summaries written into the
model directory, and biweekly milestone meetings with the
supervisor at which the artefacts produced in the previous
sprint are reviewed and the priorities for the next one are
agreed.

% ─────────────────────────────────────────────────────────────────────────────
\section{Project Planning}
\label{sec:mgmt-planning}

\subsection{Work-breakdown structure}

The plan spans 21 weeks (February--June 2026), organised into five
sprint-aligned work packages (WP1--WP5) plus a transversal
management work package (WP0) that runs in parallel throughout the
project.
Execution control uses \textbf{fixed four-week sprints} aligned
with the work packages (each WP1--WP4 is one sprint; WP5 closes
the project over five weeks).
Supervisor meetings keep a \textbf{biweekly} cadence within this
structure: each sprint holds two meetings, a mid-sprint check-in at
its midpoint and an end-of-sprint review at its boundary (the
reviews are the milestones M1--M4 in \Cref{fig:gantt}, with the
final defence marked as M5).
Over the 21 weeks this amounts to roughly ten meetings, the activity
budgeted as T0.5 (25\,h) in \Cref{tab:task-summary} and
referred to elsewhere simply as the biweekly milestone meetings.
This four-week grouping is a deliberate revision of the two-week
sprint cadence in the initial GEP plan: aligning each sprint with a
single work package gave the train--evaluate--improve loop a coherent
deliverable per sprint and accommodated the late addition of the
chord-recognition track (detailed in the WP4 and WP5 descriptions
below) without fragmenting the schedule.
Backlog priorities, risks, and effort estimates are reviewed at
every sprint boundary and may be revised in either direction.

The six work packages have the following internal logic.

\textit{WP0 -- Project management (130\,h, concurrent).}
This package contains the activities that span the project rather
than belonging to any one technical phase: the writing of the GEP
deliverables, the biweekly supervisor meetings, and the
continuous writing of the thesis document.
It is intentionally allocated a substantial budget (\SI{26}{\percent}
of total effort) because the GEP coursework requires four
deliverables in its own right and because deferring thesis
writing until the final weeks is one of the principal failure
modes of single-person projects.

\textit{WP1 -- Literature review and baseline (78\,h).}
This package establishes the analytical foundation: a literature
survey of OMR (the material that becomes \Cref{ch:soa}),
an exploration of the public OMR datasets that informs the
data-strategy choice, and a classical computer-vision baseline
that provides a quantitative reference against which the
CRNN--CTC results can be compared.
The classical baseline also serves a risk-control function: if
the CRNN--CTC model fails to converge, the baseline ensures the
project still has measurable results to report.

\textit{WP2 -- Dataset construction (73\,h).}
This package builds the synthetic training corpus that the
CRNN--CTC requires.
The output of WP2 is the LMX vocabulary, the LilyJAZZ rendering
pipeline, the scan-simulation augmentation pipeline, and the
train/validation/test splits.
WP2 must complete before WP3 can begin in earnest, because the
model architecture is dimensioned to the vocabulary the dataset
produces.

\textit{WP3 -- CRNN--CTC development (75\,h).}
This package designs and implements the model, the training loop,
and the evaluation framework.
The training-pipeline task (T3.2) overlaps deliberately with the
evaluation-framework task (T3.4), since the latter can be
developed against synthetic CRNN outputs before the former has
fully converged.
The package closes with an initial end-to-end training run whose
outputs feed directly into WP4.

\textit{WP4 -- Iterative refinement (68\,h).}
This package is the primary site of the train--evaluate--improve
loop (\Cref{sec:mgmt-methodology}); its 68-hour budget
covers two internal iterations within Sprint 4 (weeks 13--16),
each closing one pass of refinement and ablation.
The chord-recognition track (synthetic strip rendering, the chord
CRNN training and fine-tuning runs, and the chord post-processing
filter) is absorbed into T4.2 and T4.3 as a parallel branch of
the same refinement loop rather than as a separate work package,
which is why the WP4 budget is not enlarged despite the additional
scope.

\textit{WP5 -- Final evaluation and defence (85\,h).}
This package covers the final evaluation on held-out data, the
qualitative error analysis, a conditional extension task (T5.3,
budgeted at 20\,h and contingent on the monophonic objective
having been achieved), the final integration pass on the thesis
document, and the defence preparation.
The conditional extension task is the most important schedule
buffer in the plan, since it can be reduced to zero without
compromising the core deliverable.
T5.3 is allocated to fine-tuning the chord CRNN against
hand-labelled real Real Book strips (using the
\code{extract\_real\_chord\_strips} utility) and to tightening the
chord post-processing grammar, which together realise the
chord-recognition extension foreshadowed in
\Cref{sec:scope}.

\subsection{Tasks, dependencies, and resources}

The plan totals 509\,h across 31 tasks, distributed over the
21-week schedule at an average of approximately 24\,h per week;
the full per-task breakdown with effort estimates, dependency
chain, and principal resources is given in \Cref{tab:task-summary}.
Strict dependencies require a predecessor to be complete before
the successor can start; partial dependencies (marked
``\textit{partial}'') require only sufficient maturity.
Concurrency is limited to naturally compatible pairs, such as
developing the evaluation framework (T3.4) while the initial
training run (T3.3) executes on the GPU.

The plan was deliberately scoped to 509\,h rather than to the
Art.~17 nominal envelope of $18\,\text{ECTS} \times 30\,\text{h/ECTS}
= 540\,\text{h}$; the resulting 31-hour margin is held as an
undistributed reserve and is not committed to any specific work
package.
The accounting in \Cref{tab:hr-budget} and the budget total of
\Cref{tab:total-budget} are computed against the 509-hour
baseline.

\subsection{Gantt chart}

\Cref{fig:gantt} shows the full 21-week timeline, grouped by work
package, with sprint-review milestones M1--M4 and the final defence
(M5).

\begin{landscape}
\begin{figure}[p]
    \centering
    \begin{ganttchart}[
        hgrid,
        vgrid={dotted},
        x unit=0.84cm,
        y unit title=0.5cm,
        y unit chart=0.35cm,
        title height=1,
        title label font=\scriptsize,
        bar label font=\tiny,
        bar height=0.6,
        group label font=\tiny\bfseries,
        milestone label font=\tiny\bfseries,
    ]{1}{21}
        \gantttitle{February}{4}
        \gantttitle{March}{4}
        \gantttitle{April}{5}
        \gantttitle{May}{4}
        \gantttitle{June}{4} \\
        \gantttitlelist{1,...,21}{1} \\

        \ganttgroup[group/.append style={fill=gpMgmt!30}]{WP0: Project Management}{1}{21} \\
        \ganttbar[bar/.append style={fill=gpMgmt!60}]{T0.1 Context and scope}{1}{3} \\
        \ganttbar[bar/.append style={fill=gpMgmt!60}]{T0.2 Planning}{3}{5} \\
        \ganttbar[bar/.append style={fill=gpMgmt!60}]{T0.3 Budget and sustain.}{5}{7} \\
        \ganttbar[bar/.append style={fill=gpMgmt!60}]{T0.4 GEP final}{7}{9} \\
        \ganttbar[bar/.append style={fill=gpMgmt!40}]{T0.5 Meetings}{1}{21} \\
        \ganttbar[bar/.append style={fill=gpMgmt!40}]{T0.6 Thesis writing}{5}{21} \\

        \ganttgroup[group/.append style={fill=gpSprint1!30}]{WP1: Sprint 1}{1}{4} \\
        \ganttbar[bar/.append style={fill=gpSprint1!60}]{T1.1 Literature survey}{1}{2} \\
        \ganttbar[bar/.append style={fill=gpSprint1!60}]{T1.2 Dataset exploration}{2}{3} \\
        \ganttbar[bar/.append style={fill=gpSprint1!60}]{T1.3 CV baseline}{2}{4} \\
        \ganttbar[bar/.append style={fill=gpSprint1!60}]{T1.4 Baseline eval.}{3}{4} \\
        \ganttmilestone[milestone/.append style={fill=gpMilestone}]{M1 review}{4} \\

        \ganttgroup[group/.append style={fill=gpSprint2!30}]{WP2: Sprint 2}{5}{8} \\
        \ganttbar[bar/.append style={fill=gpSprint2!60}]{T2.1 Encoding}{5}{5} \\
        \ganttbar[bar/.append style={fill=gpSprint2!60}]{T2.2 Rendering}{5}{7} \\
        \ganttbar[bar/.append style={fill=gpSprint2!60}]{T2.3 Augmentation}{7}{7} \\
        \ganttbar[bar/.append style={fill=gpSprint2!60}]{T2.4 Validation}{7}{8} \\
        \ganttmilestone[milestone/.append style={fill=gpMilestone}]{M2 review}{8} \\

        \ganttgroup[group/.append style={fill=gpSprint3!30}]{WP3: Sprint 3}{9}{12} \\
        \ganttbar[bar/.append style={fill=gpSprint3!60}]{T3.1 Architecture}{9}{10} \\
        \ganttbar[bar/.append style={fill=gpSprint3!60}]{T3.2 Training pipeline}{10}{11} \\
        \ganttbar[bar/.append style={fill=gpSprint3!60}]{T3.3 Initial training}{11}{12} \\
        \ganttbar[bar/.append style={fill=gpSprint3!60}]{T3.4 Evaluation framework}{11}{12} \\
        \ganttmilestone[milestone/.append style={fill=gpMilestone}]{M3 review}{12} \\

        \ganttgroup[group/.append style={fill=gpSprint4!30}]{WP4: Sprint 4}{13}{16} \\
        \ganttbar[bar/.append style={fill=gpSprint4!60}]{T4.1 Error analysis}{13}{13} \\
        \ganttbar[bar/.append style={fill=gpSprint4!60}]{T4.2 Data fixes}{13}{14} \\
        \ganttbar[bar/.append style={fill=gpSprint4!60}]{T4.3 Refinement}{14}{15} \\
        \ganttbar[bar/.append style={fill=gpSprint4!60}]{T4.4 Ablations}{15}{16} \\
        \ganttmilestone[milestone/.append style={fill=gpMilestone}]{M4 review}{16} \\

        \ganttgroup[group/.append style={fill=gpSprint5!30}]{WP5: Sprint 5}{17}{21} \\
        \ganttbar[bar/.append style={fill=gpSprint5!60}]{T5.1 Final eval.}{17}{17} \\
        \ganttbar[bar/.append style={fill=gpSprint5!60}]{T5.2 Error analysis}{17}{18} \\
        \ganttbar[bar/.append style={fill=gpSprint5!60}]{T5.3 Extension (cond.)}{17}{19} \\
        \ganttbar[bar/.append style={fill=gpSprint5!60}]{T5.4 Thesis final}{19}{20} \\
        \ganttbar[bar/.append style={fill=gpSprint5!60}]{T5.5 Defence prep.}{20}{21} \\
        \ganttmilestone[milestone/.append style={fill=gpMilestone}]{M5 defence}{21} \\
    \end{ganttchart}
    \caption{Gantt chart for the 21-week project (February--June 2026),
    grouped by work package, with sprint-review milestones M1--M4 and the
    final defence (M5).}
    \label{fig:gantt}
\end{figure}
\end{landscape}

% ─────────────────────────────────────────────────────────────────────────────
\section{Risk Register and Contingency Plans}
\label{sec:mgmt-risks}

The risk register identifies six principal risks that were
foreseeable at the start of the project and for which explicit
mitigations or alternative plans were prepared in advance.
Each risk is rated on two axes, likelihood and impact, using a
three-level scale (low, medium, high), and is paired with an
alternative plan, an estimate of how much schedule the activation
of that plan would consume, and an estimate of any additional
resources it would require.
The full register, including the alternative plans and their
schedule and resource costs, is given in \Cref{tab:risk-register}.

Each entry also has a \emph{detection trigger}, the observable
signal that tells the project the risk has materialised:
\begin{itemize}
    \item \textbf{R1 (ground-truth data):} labelling throughput below
    roughly ten usable staves per hour in the first session, which
    would push the annotation set past its allotted window.
    \item \textbf{R2 (staff detection):} the detector missing or
    mis-segmenting more than about one staff in ten on a spot-check
    set of real scanned pages.
    \item \textbf{R3 (encoding design):} a non-trivial fraction of
    dataset tokens mapping to \code{<unk>}, or the vocabulary needing
    structural revision after the first training run.
    \item \textbf{R4 (underfitting / CTC misalignment):} validation
    SER plateauing above the \SI{2}{\percent} clean-split target (or
    the CTC loss failing to descend) after the first ten-to-fifteen
    epochs, or \code{zero\_infinity} silently masking samples whose
    CTC input length falls below their label length.
    \item \textbf{R5 (scope creep toward polyphony):} any
    implementation effort logged against polyphony, or T5.3 scope
    expanding beyond a feasibility discussion at a sprint review.
    \item \textbf{R6 (compute constraints):} a required training
    configuration exceeding the \SI{12}{\giga\byte} VRAM budget at the
    minimum viable batch size, or workstation contention stretching a
    single training run beyond about one day.
\end{itemize}

The plan also carries explicit schedule slack: conservative
estimation in the two highest-uncertainty tasks (dataset rendering
and model refinement), progressive thesis writing throughout
execution, and the conditional extension task T5.3, which functions
as the principal buffer and can be reduced to zero without
compromising the core objective (\Cref{sec:mgmt-planning}).
Together these absorb the schedule consequences of the lower-impact
risks without triggering the budget-control mechanism of
\Cref{sec:mgmt-control}.

\begin{table}[ht]
    \centering
    \caption{Risk register with alternative plans and estimated
    impact. Likelihood and impact are rated on a three-level scale:
    Low, Med(ium), or High.}
    \label{tab:risk-register}
    \scriptsize
    \setlength{\tabcolsep}{4pt}
    \renewcommand{\arraystretch}{1.35}
    \begin{tabularx}{\linewidth}{@{} c >{\raggedright\arraybackslash}p{2.2cm} >{\centering\arraybackslash}p{1.4cm} >{\centering\arraybackslash}p{1.4cm} >{\raggedright\arraybackslash}p{2.8cm} >{\raggedright\arraybackslash}p{2.2cm} >{\raggedright\arraybackslash}p{1.6cm} @{}}
        \toprule
        \textbf{ID} & \textbf{Risk} & \textbf{Likelihood} & \textbf{Impact} & \textbf{Alternative plan} & \textbf{Schedule impact} & \textbf{Extra resources} \\
        \midrule
        R1 & Limited ground-truth data on real Real Book pages
            & High & High
            & Fall back to synthetic / typeset sources and strengthen the augmentation pipeline.
            & Reallocate up to 10\,h.
            & None. \\ \tfgrowsep
        R2 & Staff-detection errors propagating downstream
            & Med & Med
            & Use manually cropped staff regions for critical evaluations.
            & Shift hours from T5.3 to T5.4.
            & None. \\ \tfgrowsep
        R3 & Output-encoding design complexity (LMX vocabulary)
            & Med & Med
            & Start from a minimal vocabulary and extend incrementally after stable runs.
            & Up to 6\,h redesign.
            & None. \\ \tfgrowsep
        R4 & Model underfitting or CTC misalignment
            & Med & High
            & Simplify vocabulary, retrain with cleaned subsets, adjust image scale.
            & Up to 10\,h from buffer.
            & Additional GPU time. \\ \tfgrowsep
        R5 & Scope creep toward polyphony
            & Med & High
            & Restrict extension to feasibility discussion only.
            & Recover up to 18\,h.
            & None. \\ \tfgrowsep
        R6 & Computational resource constraints
            & Low/Med & Med
            & Reduce model/data size or rent a temporary cloud-GPU window.
            & Up to 5\,h setup.
            & Cloud GPU rental (see \Cref{tab:incidentals}). \\
        \bottomrule
    \end{tabularx}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Economic Management}
\label{sec:mgmt-budget}

\subsection{Budget structure}

The budget follows the standard software-project cost framework
taught in the FIB GEP module~\cite{gepModule24EconomicManagement}
and is organised into four components: \textit{human resources by
activity}, itemised against the task structure of
\Cref{sec:mgmt-planning}; \textit{general costs}, covering
hardware amortisation, electricity and connectivity overheads, and
software-licence costs (zero, since the entire toolchain is open
source); \textit{incidentals}, which translate each risk's
contingency action into an expected-value contribution (full cost
$\times$ likelihood); and a \textit{contingency reserve} of fifteen
per cent of the baseline (human resources plus general costs) for
the residual uncertainty that the risk register cannot enumerate
explicitly.
The wider sustainability framing of this budget (useful life,
break-even, and the economic-impact assessment proper) is
deferred to \Cref{sec:sust-economic}, which uses the
totals derived here as its inputs.

\subsection{Human resources}

Hourly rates are derived from the Spanish ICT salary survey
INE-EAES~\cite{ineEncuestaEstructuraSalarial2023} and loaded with
a \num{1.35} multiplier that captures the employer's
social-security contributions in the General Regime of the
Spanish social-security
system~\cite{segSocialCotizacion2025}, then converted to an hourly
figure assuming a working year of 1\,760 hours, as prescribed by the
FIB GEP economic-management
module~\cite{gepModule24EconomicManagement}.
Because the project is executed by a single person carrying out all
four roles, each task in \Cref{tab:task-summary} is costed at the
loaded rate of the role that predominantly performs it.
The full per-role rate derivation and the project-weighted blended
rate are given in \Cref{app:budget-detail}; the resulting
\EUR{}13\,615.57 human-resources total is distributed across work
packages in \Cref{tab:hr-budget}.
This figure dominates the budget: roughly \SI{85}{\percent} of
the project cost is engineering time, the expected profile for a
research-intensive software project where tooling and hardware are
commodity items.

\begin{table}[ht]
    \centering
    \caption{Human-resources budget by work package.}
    \label{tab:hr-budget}
    \small
    \setlength{\tabcolsep}{5pt}
    \renewcommand{\arraystretch}{1.35}
    \begin{tabularx}{\linewidth}{@{} c L r r @{}}
        \toprule
        \textbf{WP} & \textbf{Work package} & \textbf{Hours} & \textbf{Cost (\EUR{})} \\
        \midrule
        WP0 & Project management              & 130 & 3\,482.50 \\ \tfgrowsep
        WP1 & Literature review and baseline  & 78  & 2\,066.20 \\ \tfgrowsep
        WP2 & Dataset construction            & 73  & 1\,992.55 \\ \tfgrowsep
        WP3 & CRNN--CTC development           & 75  & 2\,075.37 \\ \tfgrowsep
        WP4 & Iterative refinement            & 68  & 1\,859.10 \\ \tfgrowsep
        WP5 & Final evaluation and defence    & 85  & 2\,139.85 \\
        \midrule
        & \textbf{TOTAL human resources}      & \textbf{509} & \textbf{13\,615.57} \\
        \bottomrule
    \end{tabularx}
\end{table}

\subsection{General costs, contingency, and incidentals}

\textit{General costs} (\Cref{tab:general-costs}) cover the
share of fixed assets and overheads attributable to the 21 weeks
of the project: hardware amortisation on the workstation, monitor,
and peripherals at a four-year linear rate prorated for the
project duration (\EUR{}125.69); software licences (zero, since the
entire toolchain, Python, PyTorch, OpenCV, LilyPond, PyMuPDF,
albumentations, FastAPI, and Poetry, is open source); and overheads
covering electricity and internet connectivity, derived from the
workstation's average power draw and the Spanish domestic-tariff
rate~\cite{reeInformeSistemaElectrico2024} on the assumption that
the workstation is dedicated to the thesis for an average of
twenty hours per week.
The hardware valuation and the amortisation arithmetic are detailed
in \Cref{app:budget-detail}.

\textit{Contingency reserve.}
A contingency reserve of \SI{15}{\percent} is applied over the
sum of human resources and general costs:
\[
\text{Contingency} = 0.15 \times (\EUR{}13\,615.57 + \EUR{}250.69)
                   = \EUR{}2\,079.94 .
\]
The fifteen per cent figure is the project-management default
for software projects of this size and follows the guidance of
the FIB GEP economic-management
module~\cite{gepModule24EconomicManagement}.
It is not an estimate of any specific risk but a residual buffer
for the unknown-unknown category: the cost overruns that the
risk register cannot anticipate by definition.

\textit{Incidentals.}
The incidentals figure of \EUR{}150.96
(\Cref{tab:incidentals}) is the expected-value sum of the three
cost-bearing entries of the risk register---R1, R4, and R6
(\Cref{sec:mgmt-risks}).
R5 is listed in the same table at zero cost for traceability with the
risk register: its contingency is a no-cost reallocation of planned
effort, so it contributes nothing to the total but is kept visible
alongside the priced entries.
Each entry is the full cost of the priced contingency action
multiplied by the probability that this specific
action, rather than a no-cost effort reallocation, is required.
These activation probabilities are therefore lower than the bare
likelihood ratings of \Cref{tab:risk-register}: R1, for
instance, is rated likelihood-high, but the priced cleaning
cycle is triggered only when augmentation tuning alone is
insufficient, an outcome estimated at roughly one in four
materialisations.
The total is small in absolute terms because most of the
register's contingency actions consist of reallocating planned
effort within the human-resources budget rather than acquiring
new resources; only the cloud-GPU rental and the additional
data-cleaning cycles incur direct monetary cost.

\begin{table}[ht]
    \centering
    \caption{General costs.}
    \label{tab:general-costs}
    \small
    \setlength{\tabcolsep}{5pt}
    \renewcommand{\arraystretch}{1.35}
    \begin{tabularx}{\linewidth}{@{} L r @{}}
        \toprule
        \textbf{Item} & \textbf{Cost (\EUR{})} \\
        \midrule
        Hardware amortisation (desktop, monitor, peripherals)  & 125.69 \\ \tfgrowsep
        Software licences (open-source stack)                  & 0.00 \\ \tfgrowsep
        Overheads (electricity + internet)                     & 125.00 \\
        \midrule
        \textbf{TOTAL general costs}                           & \textbf{250.69} \\
        \bottomrule
    \end{tabularx}
\end{table}

\begin{table}[H]
    \centering
    \caption{Incidentals derived from the risk register.
    Each entry is the cost of the priced contingency action
    weighted by the probability that this specific action
    (rather than a no-cost effort reallocation) is required;
    these activation probabilities are lower than the bare
    likelihood ratings of \Cref{tab:risk-register}.}
    \label{tab:incidentals}
    \small
    \setlength{\tabcolsep}{5pt}
    \renewcommand{\arraystretch}{1.35}
    \begin{tabularx}{\linewidth}{@{} >{\raggedright\arraybackslash}p{1.0cm} L r r r @{}}
        \toprule
        \textbf{Risk} & \textbf{Contingency action} & \textbf{Full cost (\EUR{})} & \textbf{Prob.} & \textbf{Weighted (\EUR{})} \\
        \midrule
        R1 & Additional data cleaning and augmentation cycle    & 276.10 & 25\,\% & 69.03 \\ \tfgrowsep
        R4 & Retraining cycle with simplified vocabulary        & 276.10 & 20\,\% & 55.22 \\ \tfgrowsep
        R6 & Temporary cloud-GPU rental and setup               & 178.05 & 15\,\% & 26.71 \\ \tfgrowsep
        R5 & Scope reduction in extension task                  & 0.00   & 30\,\% & 0.00  \\
        \midrule
        \multicolumn{4}{r}{\textbf{TOTAL incidentals}}          & \textbf{150.96} \\
        \bottomrule
    \end{tabularx}
\end{table}

\subsection{Total budget}

\begin{table}[H]
    \centering
    \caption{Total project budget.}
    \label{tab:total-budget}
    \small
    \setlength{\tabcolsep}{5pt}
    \renewcommand{\arraystretch}{1.35}
    \begin{tabularx}{\linewidth}{@{} L r @{}}
        \toprule
        \textbf{Budget component} & \textbf{Cost (\EUR{})} \\
        \midrule
        Human resources       & 13\,615.57 \\ \tfgrowsep
        General costs         & 250.69 \\ \tfgrowsep
        Contingency (15\,\%)  & 2\,079.94 \\ \tfgrowsep
        Incidentals           & 150.96 \\
        \midrule
        \textbf{TOTAL project budget} & \textbf{16\,097.16} \\
        \bottomrule
    \end{tabularx}
\end{table}

\subsection{Budget control}
\label{sec:mgmt-control}

Budget tracking is performed at each sprint review using three
indicators drawn from the earned-value-management (EVM)
framework~\cite{pmiBOKGuide2021}: the cost variance (CV) and cost
performance index (CPI), which measure cost health, and the schedule
variance (SV), which measures progress against plan, all expressed in
the common currency of budgeted cost.
The definitions of these indicators and their underlying BCWS, BCWP,
and ACWP quantities are given in \Cref{app:evm}.

\textit{Trigger thresholds.}
Corrective actions are triggered when at least one of the
following conditions holds in a sprint review:
$|\mathrm{CV}| > \SI{10}{\percent}$ of the sprint budget,
$\mathrm{SV} < -\SI{5}{\percent}$ of the sprint budget, or
$\mathrm{CPI} < 0.90$ for two consecutive reviews.
A single low CPI reading is not by itself a trigger because
single sprints have substantial natural variance in a
single-person project; sustained underperformance over two
sprints is required before the corrective mechanism activates.
Once triggered, the corrective action is recorded in the sprint
review note and may activate one of the entries of the risk
register (\Cref{tab:risk-register}): typically the
reallocation of effort from the conditional extension task
(T5.3) to the affected work package, or the acceptance of the
short cloud-GPU rental from the incidentals budget.
