\chapter{Sustainability and Social Commitment}
\label{ch:sustain}

This chapter records the sustainability dimension of the project,
following the matrix and reflection framework set out in the FIB
GEP sustainability guide~\cite{gepSustainabilityReport2018}, in
accordance with the requirement of Article~12 of the TFG normativa
(Acord Consell de Govern UPC 07/2023).
\Cref{sec:sust-selfassessment} reports the self-assessment
that opened the project; \Cref{sec:sust-matrix} fills the
three-dimensional sustainability matrix (economic, environmental,
and social) at the project's design-and-development stage and at
its anticipated useful life; \Cref{sec:sust-commitment}
records the ethical position and open-science commitments adopted;
and \Cref{sec:sust-final} revisits the assessment in light
of the completed work.
The discussion is anchored to four UN Sustainable Development Goals
(SDGs)~\cite{gepSustainabilityReport2018} that the project plausibly
touches: SDG~4 (Quality Education), SDG~9 (Industry, Innovation and
Infrastructure), SDG~10 (Reduced Inequalities), and SDG~12
(Responsible Consumption and Production).
The relevant goal is named within each dimension below rather than
in a separate column of the matrix.
The chapter also consolidates the applicable-regulations,
intellectual-property, and personal-data analysis for the whole
thesis (\Cref{sec:sust-commitment}), which is referenced
forward from \Cref{ch:mgmt} but discussed only here.

% ─────────────────────────────────────────────────────────────────────────────
\section{Self-Assessment Summary}
\label{sec:sust-selfassessment}

The GEP sustainability questionnaire identified the author's
strongest area as the \textbf{economic} dimension of software
projects, especially cost estimation, trade-off analysis, and
management control.
The author is comfortable reasoning about whether a technical
decision is financially realistic within a constrained project, and
about how planning choices affect viability and long-term
maintenance cost.
The questionnaire also surfaced a medium level in social reflection:
accessibility and user benefit are routinely considered, but
technical performance still tends to take priority over social
impact when the project is under time pressure.

The weakest area was the \textbf{environmental} lifecycle
perspective.
At the start of the project, environmental impact was associated
mainly with direct energy consumption during model training.
The reflection process surfaced additional factors that had not
previously been analysed in depth: the hardware manufacturing
footprint, end-of-life device management, and the difference
between one-time experimentation and sustained deployment impact.
The reflection also made clear that selecting lighter architectures
is not only a computational decision but also an environmental one,
a point explicitly reflected in the architectural justification of
\Cref{sec:mgmt-justification}.

Three concrete improvement commitments emerged for the remainder
of the project.

\textit{First}, energy-related impact should be quantified wherever
possible rather than asserted qualitatively.
The energy estimate of \Cref{sec:sust-environmental} is the
result of this commitment: it is computed from the GPU hours
consumed by the training runs, multiplied by the rated power draw
of the workstation and converted to CO\textsubscript{2} using the
official Spanish emission
factor~\cite{reeInformeSistemaElectrico2024,mitecoFactorEmision2025}.
The calculation is coarse but grounded in measurable quantities
rather than in adjectives; the methodology is detailed in
\Cref{sec:sust-environmental}.

\textit{Second}, architecture choices should be justified with both
performance and efficiency criteria.
The comparison table in \Cref{sec:mgmt-justification} includes
``computational feasibility'' and ``data requirements'' as
first-class criteria alongside accuracy, and the backbone selected
in \Cref{sec:design-model} carries the environmental dividends
quantified in \Cref{sec:sust-environmental}.

\textit{Third}, the discussion of who benefits from the resulting
tool, and under what limitations, must be explicit rather than
implicit.
This commitment shapes the social-dimension discussion in
\Cref{sec:sust-social} and the ethical-position statement
in \Cref{sec:sust-commitment}; both name specific beneficiary
groups (musicians, educators, archivists, researchers) and specific
limitations (monophonic contract, treble-clef-only inputs, residual
SER profile) rather than appealing to vague ``users'' and ``edge
cases''.

The questionnaire transformed sustainability from a formal reporting
requirement into a design constraint that guides technical decisions
throughout the project lifecycle.
It also reframed project quality as a balance between technical
success, economic feasibility, environmental responsibility, and
social value, with each of those four facets reflected somewhere in
the project plan rather than treated as a downstream addendum.

% ─────────────────────────────────────────────────────────────────────────────
\section{Sustainability Matrix}
\label{sec:sust-matrix}

\Cref{tab:sustainability-matrix} summarises the
three-dimensional sustainability matrix as used in the GEP
framework.
The remainder of this section discusses each dimension in turn,
identifying the principal cost or benefit (PPP, ``the project as
designed and developed'') and the useful-life perspective (the
impact once the system is in service).

\begin{table}[h!]
    \centering
    \caption{Three-dimensional sustainability matrix for the project, contrasting the design-and-development stage (PPP) against the deployment and useful-life perspective.}
    \label{tab:sustainability-matrix}
    \small
    \setlength{\tabcolsep}{5pt}
    \renewcommand{\arraystretch}{1.35}
    \begin{tabularx}{\linewidth}{@{} >{\raggedright\arraybackslash}p{2.3cm} L L @{}}
        \toprule
        \textbf{Dimension} & \textbf{PPP (design and development)} & \textbf{Useful life (deployment and impact)} \\
        \midrule
        Economic
            & Total project budget of \EUR{}16\,097.16, with explicit contingency and incidentals.
              The main cost driver is specialised engineering time.
            & Open-source release and fast local inference can reduce recurrent transcription cost compared with manual copying and commercial OMR tools. \\ \tfgrowsep
        Environmental
            & Estimated energy use of $\approx 57.5$\,kWh for the full project, mitigated by mixed-precision training, early stopping, and the reuse of existing
              hardware~\cite{reeInformeSistemaElectrico2024,mitecoFactorEmision2025}.
            & One-time training cost is amortised over repeated inference; local execution reduces cloud dependency for common use cases. \\ \tfgrowsep
        Social
            & The project strengthens the author's personal technical competencies and promotes responsible data and software practices.
            & Potential beneficiaries include musicians, educators, archivists, and researchers through improved accessibility, searchability, and preservation of music documents. \\
        \bottomrule
    \end{tabularx}
\end{table}

\subsection{Economic dimension}
\label{sec:sust-economic}

\textbf{PPP (design and development).}
The total project cost (\EUR{}16\,097.16) is constructed in detail
in \Cref{sec:mgmt-budget}; what matters for the sustainability
reading is the structure of that budget rather than the headline
number.
The expenditure is dominated by specialised engineering time, a
profile typical of research-intensive software projects in which
labour outweighs material amortisation, so reductions in scope or
schedule translate directly into reductions in cost.
The interpretation relevant to SDG~9 is that the project's economic
footprint is not material-intensive: no new physical infrastructure
is built, and the budget is mostly a transfer of human capital.

\textbf{Useful life and state of the art.}
The alternatives for lead-sheet digitisation are manual
transcription and general-purpose OMR tools, and the project's
economic improvement must be evaluated against both.
Manual transcription is expensive in recurrent cost (a trained
musician's time, at conservatory or freelance rates) and in
turnaround time (typically of the order of minutes to tens of
minutes per page, depending on density and clarity).
General-purpose OMR tools, both commercial (SmartScore, PhotoScore,
SharpEye) and open-source (Audiveris), reduce the recurrent cost
but require substantial correction effort on the jazz-specific
notation studied here, because their internal heuristics are tuned
for classical engraving styles (\Cref{sec:soa-classical}).
A specialised CRNN--CTC model targeted at the Real Book engraving
style reduces the correction workload further, on the class of
input it was trained for, and eliminates the licence costs that
the commercial alternatives charge per seat.

\textbf{Improvement contribution.}
The principal economic improvement is not zero-cost digitisation
but a \textit{lower cost per corrected page} after deployment.
A user of the system still has to verify and correct the output
manually for any application that depends on note-perfect
transcription, but the per-page correction time is reduced
because the model handles the bulk of the symbol recognition
correctly (\Cref{ch:results}).
A second improvement is research reproducibility: the open-source
release of the code, the training scripts, and the trained model
weights lowers the entry cost for future OMR research targeting
the same or adjacent domains, because the synthetic data pipeline
and the CRNN--CTC training stack do not have to be rebuilt from
scratch.

\subsection{Environmental dimension}
\label{sec:sust-environmental}
This dimension is the one most directly aligned with SDG~12
(Responsible Consumption and Production) and is treated quantitatively.

\textbf{PPP (design and development).}
The project's estimated energy consumption is approximately
\SI{57.5}{\kilo\watt\hour}, derived from a coarse but reproducible
multiplication of the rated power draw of the development
workstation by the cumulative GPU-engaged time over the twenty-one
week schedule.
The workstation pairs an NVIDIA RTX~3060 GPU (\SI{170}{\watt} TDP)
with a desktop platform whose full-system draw under training load
is approximately \SI{250}{\watt}.
Cumulative GPU-engaged time totals approximately
\SI{230}{\hour}, dominated by the main CRNN training (37 converged
epochs, plus ablation and chord-OCR fine-tuning runs), with smaller
contributions from synthetic-data rendering and evaluation passes.
The energy estimate follows directly:
$\SI{230}{\hour} \times \SI{0.25}{\kilo\watt} \approx
\SI{57.5}{\kilo\watt\hour}$.\todo{cite: ML carbon-accounting methodology (Strubell et al.\ 2019, Patterson et al.\ 2021, Lacoste et al.\ 2019) for the kWh $\to$ CO\textsubscript{2} conversion convention used here.}
Applying the Spanish electricity emission factor of approximately
\SI{0.23}{\kilo\gram\per\kilo\watt\hour} of CO\textsubscript{2}
reported by Red El\'ectrica de Espa\~na and
MITECO~\cite{reeInformeSistemaElectrico2024,mitecoFactorEmision2025}
yields roughly \SI{13.2}{\kilo\gram} of CO\textsubscript{2}e
emitted during the project's execution.

The dominant contributor to the total is GPU training time.
Inference, dataset generation, and development sessions away from
the GPU contribute much less per hour because they do not engage
the high-power-draw components of the workstation.
This concentration of environmental impact in a single activity
makes it tractable: optimising training has substantially more
environmental leverage than optimising the rest of the workflow.

\textbf{Minimisation plan.}
Four mitigations together keep the training-time energy cost
within the estimated envelope.
First, the architecture choice (\Cref{sec:design-model})
favours a \textit{lightweight} backbone, ResNet18, over the
heavier alternatives that would yield only marginal accuracy
gains on monophonic input;
\Cref{sec:mgmt-justification} makes this trade-off
explicit.
Second, \textit{mixed-precision training} via PyTorch's
\code{torch.amp} reduces both VRAM occupancy and the cycle time
per training step, which translates directly into fewer
GPU-hours per converged model.
Third, \textit{early stopping} with a twelve-epoch patience halts
training once the validation SER stops improving, avoiding the
common pattern of running for a fixed number of epochs ``to be
safe''.
Fourth, the project \textit{reuses existing hardware} rather than
provisioning new equipment specifically for the thesis; the
embodied energy of manufacturing a new workstation would dwarf
the operational energy of the project by an order of magnitude.

Data generation and training are also organised in batches to
avoid redundant runs: the dataset is fixed before model training
begins, so that data-pipeline iteration and model iteration do
not compound in cost.
Each significant training run is logged and its CSV summary
retained, which makes it possible to identify and avoid
duplicate experiments in later sprints.

\textbf{Useful life and improvement contribution.}
Once trained, inference has a low marginal energy cost compared
with retraining: a single page transcription completes in
seconds on consumer hardware and consumes a small fraction of a
watt-hour of energy.
The energy break-even point of the project, i.e.\ the number of
inference invocations at which the operational energy of inference
equals the one-time training cost, is therefore high enough that
the trained model can be regarded as an amortised asset.
Sharing the trained weights and the training scripts spreads the
one-time training cost across multiple downstream users and
multiple academic iterations, so that the per-application share
of the project's environmental footprint shrinks as the model is
re-used.

\textbf{Embodied and end-of-life impact.}
The operational figure above excludes two lifecycle stages that
are harder to quantify but worth naming explicitly.
The \textit{embodied} energy of manufacturing the GPU and the rest
of the workstation is paid once at fabrication, amortises across
every project run on the same hardware, and is the main reason the
project deliberately reuses an existing workstation rather than
provisioning new equipment.
At \textit{end of life}, the same GPU is expected to enter the
domestic e-waste channel rather than be disposed of as bulk waste,
and the trained-weight artefacts incur a small archival storage
cost that is dwarfed by the cost of \emph{retraining} a successor
model once architectures, datasets, or notation styles drift; the
project mitigates this by publishing training scripts alongside the
weights so a replacement can be re-derived rather than learned from
scratch.\todo{cite: figure for embodied carbon of a consumer GPU or workstation if a defensible source can be located (e.g.\ NVIDIA or Dell product carbon-footprint disclosures); otherwise leave as qualitative.}

\subsection{Social dimension}
\label{sec:sust-social}
The social dimension touches SDG~4 (Quality Education) through
improved access to editable teaching materials, and SDG~10
(Reduced Inequalities) through the accessibility argument developed
below.

\textbf{PPP (design and development).}
In personal-development terms the project strengthens competencies
in three areas relevant to the author's future professional work:
machine-learning engineering (architecture design, training-loop
implementation, evaluation methodology), project planning
(work-breakdown structures, sprint cadence, budget management), and
scientific communication (technical writing, figure-driven argument,
and the citation discipline required by a thesis-length document).
The project also reinforces ethical awareness about the limitations
and failure modes of automated transcription: working through the
residual-error analysis of \Cref{ch:results} produces a more
concrete picture of what the model gets right and wrong than a
single aggregate metric can convey.

\textbf{Useful life and beneficiaries.}
There is real demand for digitising lead sheets used in music
education and performance, and the system targets a notation style
(Real Book engraving) that is widely used and poorly served by
existing tools.
For \textit{musicians}, the downstream applications include
automatic transposition to a different key (essential for B$\flat$,
E$\flat$, and F transposing instruments), MIDI playback for
practice and accompaniment, and integration with practice apps that
depend on a symbolic representation of the score.
For \textit{teachers}, the value is in producing digital materials
that can be edited, transposed, and annotated for specific
pedagogical purposes without re-typesetting the entire piece by
hand.
For \textit{archivists}, OMR is a precondition for making large
paper collections searchable: a scanned PDF is preserved but not
discoverable, whereas a symbolic representation can be queried
melodically, harmonically, or rhythmically.
For \textit{researchers}, especially in music information retrieval
and computational musicology, a symbolic version of a large jazz
corpus enables statistical and structural analyses that cannot be
performed on raster images.

A related accessibility argument applies to \emph{sight-impaired
musicians}: the structured JSON output is in principle consumable
by screen readers and braille-music tooling, whereas a scanned PDF
is not.
Realising this benefit in practice depends on downstream software
the project does not itself ship, so the contribution is at the
level of producing a machine-readable representation rather than a
finished accessibility tool.

\textbf{Users not served.}
The narrow scope of the domain specification leaves several user
groups outside the system's contract.
Users of \emph{polyphonic} notation (piano scores, choral parts,
guitar tablature, ensemble full scores) are excluded by the
monophonic-only filter (\Cref{sec:design-data}).
Users of \emph{non-Western notation systems} (Indian sargam,
Byzantine neumes, jianpu numbered notation) are excluded entirely:
the vocabulary is built around Common Western Music Notation.
Users of \emph{handwritten} or \emph{historical engraved sources}
are excluded by the choice of training corpus (synthetic LilyJAZZ
renders of PrIMuS), which captures neither the variance of human
handwriting nor the typography of pre-20th-century editions.
Naming these exclusions is not a defect catalogue; it makes the
boundary of the social benefit explicit so that the system is not
deployed as a general-purpose music recogniser by accident.

Across the served groups, the project may improve quality of life
in two related ways: by reducing the repetitive manual work
required to convert paper notation to a machine-readable form, and
by facilitating the long-term preservation of music content that
currently exists only as paper or low-resolution scanned PDF.

\textbf{Boundaries.}
Social value depends on transparent communication of the system's
current limitations.
Four limitations are particularly important to name explicitly.
The \textit{monophonic-only contract} (\Cref{sec:design-data})
means the model is not designed to transcribe piano scores, choral
parts, or any input with more than one simultaneous voice on the
staff.
The \textit{Real-Book-specific engraving style} means the model's
accuracy on other engraving styles, such as classical Sibelius
output, historical editions, or handwritten manuscripts, is not
the accuracy reported in \Cref{ch:results} and would have to be
measured separately.
The \textit{treble-clef-only assumption} excludes any input on F
or C clefs, which is acceptable for lead sheets but disqualifying
for full scores.
The \textit{residual error rate} reported in \Cref{ch:results},
while small in aggregate, is not zero; applications that depend on
note-perfect transcription must include a human verification step.
Overpromising the system in scenarios outside the validated scope
would erode rather than enhance its social value, and would also
expose the system's users to downstream costs they have not been
warned about.

% ─────────────────────────────────────────────────────────────────────────────
\section{Commitment and Ethical Position}
\label{sec:sust-commitment}

The project's ethical position aligns with the open-science
commitments of the GEP sustainability framework, and is recorded
here as five concrete practices rather than a single statement of
principle.

\textit{Open source release.}
The source code is released under a permissive open-source
licence and is tracked in a public Git repository.
Every experiment in the thesis can be reproduced from a clean
clone using a pinned-dependency environment managed by Poetry
(\Cref{sec:impl-reproducibility}), and the trained model
weights are distributed alongside the code.
This commitment matters because OMR systems often disappear into
private repositories that cannot be inspected or extended; making
the implementation public is the most direct contribution the
project can make to the surrounding research community.

\textit{Documented evaluation protocol.}
The metric protocol used in \Cref{ch:results} is documented
in full, including the definition of the aggregate Symbol Error
Rate, the melodic SER that strips structural tokens, and the
category-level error breakdown that decomposes edits into
substitutions, insertions, and deletions per token class.
Future work can therefore replicate the present evaluation
exactly, or extend it with additional metrics, without having to
guess at how the present numbers were computed.

\textit{Licence-respecting use of third-party data.}
All third-party data used in the project is consumed in
accordance with its licence.
PrIMuS, the synthetic-supervision source, is distributed by the
Universidad de Alicante PRAIG group for non-commercial research
use~\cite{PrIMuSDataset} and is used here unmodified.
The training corpus visible to the CRNN is entirely synthetic:
LilyJAZZ renders generated from PrIMuS \texttt{.semantic} files, so
no copyrighted Real Book page is part of the music CRNN's
training-time inputs.
The chord branch is trained primarily on synthetic LilyJAZZ chord
strips, with fine-tuning on a small set of hand-labelled strips
extracted by \code{extract\_real\_chord\_strips} from privately held
Real Book scans.
The Real Book itself is a copyrighted commercial publication (Hal
Leonard), and any private scans used by the author are restricted
to evaluation and qualitative inspection in a strictly academic
context; they are never redistributed, and the released model
weights do not embed reconstructable copies of those pages, because
both CRNN-CTC pipelines learn symbolic token sequences rather than
reproducible pixel data.
This use falls within the academic exceptions granted by Spanish
intellectual-property law (\textit{Real Decreto Legislativo
1/1996}, art.~32, on quotation and illustration for teaching and
scientific research~\cite{spainLPI1996}) and by the text-and-data-mining
exception for research organisations established in art.~3 of the EU
Digital Single Market Directive (2019/790)~\cite{euDSMDirective2019}.

\textit{No personal data, no safety-critical decisions.}
The system does not process personal data at any stage: inputs are
music notation, outputs are symbolic representations of that
notation, and it is not deployed for safety-critical automated
decision-making.
The General Data Protection Regulation (GDPR,
Regulation (EU) 2016/679~\cite{euGDPR2016}) is therefore not engaged
by the project's training, inference, or deployment surface, since
no identifiable natural persons are referenced in any data stream.
This narrows the ethical surface of the project to its
intellectual-property dimension and to its honesty about its
capabilities.

\textit{Misuse considerations.}
The same open-source release that makes the system reproducible
also lowers the barrier for foreseeable misuse: at-scale digitisation
and online redistribution of copyrighted lead sheets, circumvention
of paid score libraries, or unlicensed derivative works generated
from the symbolic output.
The project mitigates these risks by limiting the released
artefacts to code, training scripts, and weights trained exclusively
on synthetic PrIMuS-derived data, without packaging any
Real-Book-derived corpus or pre-extracted symbol library.
Transparency about the system's capabilities and limitations, in
this thesis and in the public \code{README}, is the principal
non-technical mitigation: users acting in good faith are informed
about what the system does, and users acting in bad faith cannot
claim ignorance of the rights-holder relationship.

\textit{Documented limitations.}
The known limitations of the system---the monophonic contract,
the treble-clef-only assumption, the eight allowed time
signatures, and the residual SER profile---are documented
prominently in the project's \code{README}, in the project
overview documentation, and in the present thesis
(\Cref{sec:scope} and \Cref{sec:sust-social}).
Future users of the model can therefore evaluate in advance
whether the system is suitable for their use case rather than
discovering its limitations through failure in production.

% ─────────────────────────────────────────────────────────────────────────────
\section{Final Reflection}
\label{sec:sust-final}

Revisiting the sustainability assessment at the end of the project,
two updates are worth recording.

First, the \textbf{environmental} estimate of $\approx 57.5$\,kWh
made at the planning stage is consistent with the executed
training profile: the final checkpoint converged in
\num{37} epochs (\Cref{sec:res-training}) and the full
training run including ablations fits within the planned envelope.
Mixed-precision training was the single most effective mitigation:
it reduced VRAM usage enough that the entire pipeline runs on a
consumer GPU, avoiding the cloud-rental scenario reserved as the
worst-case fallback in the incidentals table
(\Cref{tab:incidentals}).

Second, the \textbf{social} value of the system depends on the
deployment surface, not just on its accuracy in isolation.
The FastAPI web service of \Cref{sec:impl-api} packages the
inference pipeline behind a single REST endpoint with no GPU
requirement on the client side, which is the deployment shape
intended to make the tool accessible to musicians, educators, and
archivists who do not have local ML expertise.
Realising that accessibility in practice requires a hosted
deployment that is outside the scope of the present thesis; what
the thesis demonstrates is that the technical preconditions for a
low-friction deployment are met.
This shape was a planned design decision, but its sustainability
significance only becomes visible once the technical work is
complete: a model with the same accuracy and no accessible
deployment surface would have substantially less potential social
impact regardless of its training-time efficiency.

A residual concern, recorded here for completeness rather than as
an unresolved issue, is the \textbf{lifecycle} of the trained model
weights themselves.
Model files have a finite useful life: they become obsolete as
new architectures are released, as datasets evolve, and as the
target notation style drifts.
Publishing the training scripts alongside the weights is the
standard mitigation: it ensures that the cost of training a
replacement model can be paid once and then redistributed, rather
than being repeated independently by each successor project.
