\chapter{Conclusions and Future Work}
\label{ch:conclusions}

This chapter closes the memoir by returning to the objectives stated in
\Cref{sec:objectives}, restating the headline quantitative results,
assessing the limitations that the work made explicit, sketching
concrete extensions that follow directly from those limitations, and
mapping the technical competences declared at TFG inscription onto the
chapters where evidence for each lives.
The intent is to give a tribunal a single place from which the rest of
the document can be audited: every claim below points to the section
where it was developed or measured.

% ─────────────────────────────────────────────────────────────────────────────
\section{Objectives Revisited}
\label{sec:conc-objectives}

The seven specific objectives listed in \Cref{sec:objectives} are
revisited below.
Each is classified as \textbf{met}, \textbf{partially met}, or
\textbf{deferred to future work}, with a one-sentence justification and
a pointer to the evidence.

\begin{itemize}
    \item \textbf{O1 — State-of-the-art survey and architectural
          justification.}
          Met.
          \Cref{ch:soa} surveys the three principal families of OMR
          systems and \Cref{sec:design-pipeline-alts,sec:design-model-alts,sec:design-twostream-alts}
          adjudicate the chosen CRNN--CTC pipeline against the alternatives.
    \item \textbf{O2 — Synthetic dataset from PrIMuS in LilyJAZZ style.}
          Met.
          The render-convert-twins-augment-vocab pipeline of
          \Cref{sec:impl-data} produces a paired
          \texttt{PNG+LMX} corpus from the original PrIMuS
          semantic files, with the engraving style aligned to the
          target domain.
    \item \textbf{O3 — Scan-simulation augmentation pipeline.}
          Met.
          The \texttt{albumentations}-based pipeline calibrated against
          real Real Book scans is described in \Cref{sec:design-data}
          and its concrete parameter set is tabulated in
          \Cref{tab:aug-pipeline}; its effectiveness is quantified by
          the small clean-to-scanned gap reported below.
    \item \textbf{O4 — CRNN--CTC model design, training, and SER
          evaluation.}
          Met.
          The model is implemented in \Cref{sec:impl-model}, trained
          with the protocol of \Cref{sec:design-training}, and
          evaluated against the acceptance criteria of
          \Cref{sec:design-domain}; the headline numbers reach
          \SI{1.19}{\percent} aggregate SER on the clean test split and
          \SI{1.28}{\percent} on the scanned split
          (\Cref{tab:results-headline}), well below the
          \SI{2}{\percent} and \SI{3}{\percent} targets set in advance.
    \item \textbf{O5 — Quantitative error analysis.}
          Met.
          The per-category breakdown of \Cref{sec:res-errors}
          decomposes residual edits into structural (barlines, ties)
          and musically substantive (pitch, octave, duration,
          accidental) classes, isolating the contribution of each
          category to the aggregate metric.
    \item \textbf{O6 — Inference pipeline with chord OCR and JSON
          output.}
          Met.
          The five-stage pipeline of \Cref{sec:design-pipeline} is
          realised in code at \Cref{sec:impl-inference}, exposed as a
          REST endpoint by the FastAPI service of
          \Cref{sec:impl-api}, and returns the JSON schema documented
          in the same section.
    \item \textbf{O7 — Domain-gap quantification and fine-tuning on
          real Real Book pages.}
          \emph{Partially met / deferred.}
          The synthetic-to-scan gap is quantified by the
          clean-versus-scanned comparison
          (\Cref{tab:results-headline}), and the fine-tuning protocol
          is specified in \Cref{sec:design-training}.
          The end-to-end measurement on hand-annotated Real Book pages
          is not concluded in this memoir and is carried forward as
          the first item of \Cref{sec:future-work}.\todo{verify:
          confirm against the final state of \S\ref{sec:res-domaingap}
          at deposit time. If real-page numbers are produced before
          deposit, upgrade this entry to ``met''. If not, the wording
          here is the honest version.}
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Key Quantitative Results}
\label{sec:conc-results}

The headline figures from \Cref{ch:results} are restated here so the
tribunal can audit the assessment above without paging back.
All numbers refer to the epoch-37 checkpoint at
\code{models/latest/best\_model.pt}, evaluated on \num{4604} held-out
samples per split with greedy CTC decoding.

\begin{itemize}
    \item \textbf{Aggregate SER:} \SI{1.19}{\percent} (clean) and
          \SI{1.28}{\percent} (scanned).
          The clean-to-scanned gap is \num{0.09} percentage points,
          well inside the one-percentage-point margin set as the
          augmentation acceptance criterion in
          \Cref{sec:design-domain}.
    \item \textbf{Melodic SER:} \SI{0.11}{\percent} (clean) and
          \SI{0.18}{\percent} (scanned), measured after stripping
          \code{measure} and tie tokens; the model gets the actual
          pitch, duration, and accidental content right on more than
          \SI{99}{\percent} of tokens.
    \item \textbf{Perfect-transcription rate:} \SI{73}{\percent}
          (clean) and \SI{71}{\percent} (scanned); roughly three out
          of four input staves come out token-identical to the ground
          truth (\Cref{sec:res-ser}).
    \item \textbf{Error composition:} approximately \SI{74}{\percent}
          of edits are \code{measure} tokens and \SI{13}{\percent} are
          ties; together these two structural classes account for
          about \SI{87}{\percent} of the residual SER, while
          per-category pitch, octave, and duration error rates each
          remain below \SI{0.15}{\percent} (\Cref{sec:res-errors}).
    \item \textbf{Beam-vs-greedy decoding:} beam search at width 5
          yields an SER improvement of at most one edit per thousand
          tokens at roughly \num{6}$\times$ the decoding cost; greedy
          decoding is therefore the deployed default
          (\Cref{sec:impl-inference}).
\end{itemize}

The four acceptance criteria from \Cref{sec:design-domain} are met
without exception on the synthetic test split.
The principal open question is the model's behaviour on genuine,
unaugmented Real Book pages, which is treated in
\Cref{sec:res-domaingap} and revisited as the leading future-work item.

% ─────────────────────────────────────────────────────────────────────────────
\section{Critical Assessment and Limitations}
\label{sec:conc-limitations}

The system meets its declared targets, but the breadth of the targets
is itself a deliberate restriction.
The list below enumerates the limitations that follow from design
decisions taken in \Cref{ch:design}, separating those that were
explicitly contracted out of scope from those that surfaced during
implementation.

\paragraph{Scope restrictions, by design.}
The target domain is monophonic treble-clef lead sheets with the
eight common time signatures of jazz practice
(\Cref{tab:domain-contract}).
Polyphony, piano grand staves, non-treble clefs, exotic meters
(\(5/8\), \(7/4\), \(11/8\), and the like), tuplets, slurs,
articulations, repeat barlines, voltas, and navigation marks
(D.C., D.S., segno, coda) are all out of scope.
None of these features appear with non-negligible frequency in the
Real Book, and extending the LMX vocabulary, the grammar fixer, and
the training distribution to cover any one of them would require
coordinated changes across the entire stack
(\Cref{sec:design-data,sec:lmx-format,sec:design-grammar}).
Each is a candidate for future work rather than a bug.

\paragraph{Synthetic-to-real domain gap.}
The model is trained entirely on PrIMuS samples re-rendered in
LilyJAZZ and augmented with a calibrated scan-simulation pipeline.
The scanned-split SER is close to the clean-split SER, which bounds
the residual gap on the augmentation distribution, but the
augmentation pipeline cannot reproduce every visual idiosyncrasy of
genuine Real Book paper: ink-pen artefacts, paper grain, show-through
from the back side of the page, and hand-written annotations are
absent from the synthetic distribution and are explicitly listed as
unimplemented in the project documentation.
The end-to-end measurement on real pages is therefore the most
important uncovered question.

\paragraph{Structural errors dominate the residual SER.}
About \SI{87}{\percent} of the residual edits are barline and tie
tokens (\Cref{sec:res-errors}).
These are largely musically irrelevant once a downstream renderer
re-bars the output, which is the reason for reporting a separate
melodic SER, but they remain the largest contributor to the headline
number and the most visible target for further improvement.
A grammar-aware decoder or a richer beat-counting validator could
reduce them further, at the cost of additional engineering.

\paragraph{No comparison to commercial OMR.}
The evaluation reports synthetic-split numbers against the
acceptance criteria of \Cref{sec:design-domain}, but does not
compare the system head-to-head against commercial OMR products
such as Audiveris or SmartScore on the same input.
A qualitative comparison on a small sample of Real Book pages would
strengthen the case for the chosen architecture and is feasible
within the existing infrastructure;
it is flagged in \Cref{sec:res-domaingap}
and noted as future work below.\todo{cite Audiveris and SmartScore if used in a comparison; placeholder citations \texttt{audiveris} and \texttt{smartscore} expected in references.bib.}

\paragraph{CTC-confidence reject gate is checkpoint-bound.}
The staff rejection gate of \Cref{sec:design-staffdetect} uses a
CTC mean log-probability threshold calibrated against the
distribution of the current checkpoint.
The threshold must be re-derived whenever the recogniser is
retrained, which is documented as a hard constraint in the
project's contributor guide.
The gate is robust within a checkpoint family but it is not
self-calibrating; a learned out-of-distribution detector would be a
cleaner long-term solution.

\paragraph{Single-page inference.}
The current pipeline processes only the first page of a multi-page
PDF (\Cref{sec:impl-inference}).
Batch multi-page handling is left to future work: it is a small
engineering task on top of the existing per-page entry point but
requires deciding how to expose multi-page output through the JSON
schema and the REST API.

\paragraph{Reading-order recovery within a system.}
Lead sheets occasionally use multi-staff systems with first- and
second-ending markings; the present pipeline detects staves
independently and does not recover their narrative reading order.
This is a layout-analysis problem that the morphological staff
detector is not designed to solve.

These limitations are honest constraints rather than failures: each
follows from a decision taken explicitly in \Cref{ch:design}, and the
acceptance criteria were set so that the system as scoped can be
adjudicated against them.

% ─────────────────────────────────────────────────────────────────────────────
\section{Future Work}
\label{sec:future-work}

Five extensions follow directly from the limitations of the previous
section.
Each is scoped against the existing codebase and infrastructure rather
than as an open research direction.

\paragraph{Real-data fine-tuning and end-to-end measurement on Real
Book pages.}
The synthetic-to-real gap remains the most important open question.
A natural follow-up is to assemble a hand-annotated set of \numrange{50}{100}
staves drawn from real Real Book scans, run the fine-tuning protocol
specified in \Cref{sec:design-training} with the convolutional
backbone frozen and only the BiLSTM and CTC head updated, and
measure aggregate and melodic SER on the same pages before and after
fine-tuning.
The infrastructure (the music-strip labelling UI of
\Cref{sec:impl-api} and the \code{--finetune-data-dir} flag of the
training CLI) is already in place;
what is missing is the manual annotation effort, which is the binding
constraint.

\paragraph{Vocabulary extension to tuplets.}
Tuplets are absent from both the training distribution and the LMX
vocabulary, so the recogniser silently flattens triplet runs to plain
notes.
Adding tuplets would require introducing paired bracket tokens
(\code{tuplet:start:N:M}, \code{tuplet:stop}) into the grammar of
\Cref{sec:lmx-format}, regenerating the vocabulary file, retraining
the model from scratch, and extending the grammar fixer to track
in-flight tuplet groups.
The Real Book uses tuplets sparingly, so the expected gain is
moderate, but it is the single largest gap in the LMX vocabulary's
coverage of the corpus.

\paragraph{Polyphonic input via stem-direction voice separation.}
The current grammar and CTC loss assume a single voice per staff
(\Cref{sec:design-model}).
Extending the system to polyphonic input would require a richer LMX
vocabulary (chord-stack tokens, voice indicators) and either a
multi-target CTC formulation or a sequence-to-sequence
decoder~\cite{riosVilaSheetMusicTransformer2024}.
A pragmatic intermediate step is to detect stem direction at the
image level, split each staff into two monophonic substrips along the
voice boundary, and run the existing recogniser on each.
This trades model complexity for a more involved preprocessor and
keeps the trained checkpoint usable, at the cost of a known failure
mode on vertically stacked noteheads whose reading order is not
recoverable from the image alone.

\paragraph{Learned staff detection and full-page layout analysis.}
The morphological staff detector of \Cref{sec:design-staffdetect}
performs well on cleanly scanned Real Book pages but is fragile on
slanted staves and partial systems
\cite{simonettaOpticalMusicRecognition2024}.
Replacing it with a learned detector (YOLO-style box
regression\todo{cite original YOLO paper, e.g.\ Redmon et al.\ 2016, in
references.bib as \texttt{redmonYouOnlyLook2016}} or a DETR-style
transformer detector) trained on a few hundred annotated pages would
unlock
multi-page batch inference and reduce the failure rate on the
hardest scans.
The current pipeline contract
(\Cref{tab:stage-contracts}) is exactly the interface the new detector
must implement, so the replacement is local: no downstream stage
needs to change.

\paragraph{Augmentation gap: paper grain and show-through.}
The augmentation pipeline does not model paper bleed-through or
realistic page-fibre texture (\Cref{sec:design-data}).
Adding these effects requires extending \code{augment\_scanned.py}
with a back-page blend (a low-opacity, horizontally-flipped second
image overlaid through a textured alpha mask) and a multi-octave
Perlin-noise paper-fibre layer applied after tone remapping.
This is a constrained engineering task within the existing
augmentation framework and would close the visible remainder of the
synthetic-to-real distribution shift on the noisier end of the
Real Book scan corpus.

% ─────────────────────────────────────────────────────────────────────────────
\section{Technical Competences Achieved}
\label{sec:conc-competences}

This section maps the GEI/Computació technical competences declared at
TFG inscription onto the chapters and sections of this memoir where
evidence for each was developed.
Both declared competences are at the in-depth level.

\subsection{CCO1.1 — Algorithmic strategies and computational
performance}

\textbf{Statement.}
``To evaluate the computational complexity of a problem, know the
algorithmic strategies which can solve it and recommend, develop and
implement the solution which guarantees the best performance according
to the established requirements.'' (In depth.)

\textbf{Where evidence lives.}
The principal venue for this competence is the comparative analysis
of architectural alternatives in \Cref{ch:design}, where every major
design decision is adjudicated against at least two rejected
alternatives on explicit complexity and performance grounds.
The pipeline decomposition of \Cref{sec:design-pipeline} and its
alternatives subsection \Cref{sec:design-pipeline-alts} weigh a
modular five-stage pipeline against monolithic full-page
transformers and tree-decoder transformers along compute-budget and
training-data axes.
The model-family decision (\Cref{sec:design-model,sec:design-model-alts})
weighs CRNN--CTC against autoregressive transformers and pure-CNN
classifiers on three axes: data efficiency, exposure-bias
amplification under length-bounded monotonic alignment, and the
fixed compute budget of a single consumer GPU.
The backbone choice (ResNet18 over deeper ResNets, EfficientNets,
ViTs, and the legacy VGG stack) is justified on parameter count,
convergence rate, and VRAM headroom for a non-trivial batch size.

The implementation chapter contributes further evidence.
The asymmetric-stride modification of the torchvision ResNet18
(\Cref{sec:impl-model}) compresses the height dimension eight times
more aggressively than the width to preserve the CTC time axis at a
\(4\times\) compression ratio, which is the smallest power-of-two
ratio at which all in-scope samples satisfy the CTC input-length
constraint.
The CTC-decoder choice and the empirical observation that beam search
at width 5 yields at most one edit per thousand tokens of improvement
at roughly \(6\times\) decode cost (\Cref{sec:impl-inference},
restated in \Cref{sec:conc-results}) is a direct
complexity-versus-accuracy trade-off measured rather than asserted.
The data-pipeline implementation (\Cref{sec:impl-data}) uses Python's
\code{multiprocessing} module for CPU-bound rendering, conversion,
and augmentation stages, since the global interpreter lock makes
threading useless for compute-bound work; the worker-pool size is a
deliberate hyperparameter rather than an accident.
The grammar fixer is implemented as a deterministic stateful walker
rather than as a constrained beam search or a weighted finite-state
transducer (\Cref{sec:design-grammar-alts}), on the explicit ground
that the grammar's finite-state size makes post-hoc enforcement
essentially free.

The acceptance criteria of \Cref{sec:design-domain} commit the
project to numeric performance targets ahead of measurement, and
\Cref{ch:results} adjudicates against them, closing the loop between
algorithmic-strategy choices and the performance they actually
deliver.

\subsection{CCO2.4 — Machine learning and extraction of information
from large data volumes}

\textbf{Statement.}
``To demonstrate knowledge and develop techniques about computational
learning; to design and implement applications and systems that use
them, including these ones dedicated to the automatic extraction of
information and knowledge from large data volumes.'' (In depth.)

\textbf{Where evidence lives.}
The full machine-learning pipeline of this thesis is the principal
evidence.
\Cref{ch:soa} surveys the three generations of OMR systems
(heuristic modular pipelines, CRNN--CTC sequence models,
transformer-based end-to-end approaches) and grounds the
architectural choice in published baselines.
\Cref{sec:design-model} commits to a CRNN--CTC stack: a ResNet18
convolutional backbone, a two-layer bidirectional LSTM, and a CTC
output head, supervised by the CTC loss on LMX token sequences.
The CTC formulation~\cite{gravesConnectionistTemporal2006} is the
mechanism by which the model learns the alignment between input
frames and output tokens without per-frame supervision; it is the
core machine-learning insight on which the entire melody-recognition
stage rests.

The data dimension is large by Bachelor's-thesis standards.
PrIMuS contributes 87\,678 monophonic incipits
(\Cref{sec:design-data}), each re-rendered in LilyJAZZ and paired
with a synthetic ``scanned'' copy through the augmentation pipeline,
yielding on the order of \num{170000} training images after
filtering.
This is the ``large data volume'' the competence statement refers
to, and the design choices that exploit it are explicit:
synthetic supervision rather than manual annotation;
sample-level oversampling of rare tokens rather than class-weighted
loss (\Cref{sec:design-training}); first-class header-less twin
samples to teach the model the continuation-line case without the
image-label desync that a training-time pixel crop would introduce;
and a deterministic split protocol
(\Cref{sec:impl-reproducibility}) that keeps validation and test
disjoint across reruns.

The structured-output side of the competence is satisfied by the
choice of LMX as the model's symbolic target rather than raw
MusicXML or a project-internal encoding
(\Cref{sec:lmx-format,sec:design-lmx-alts}): a flat
whitespace-separated token sequence that round-trips to MusicXML
through a deterministic post-processor, compact enough to serve as a
CTC target and aligned with the OLIMPIC reference
work~\cite{mayerPracticalEndToEnd2024}.
The chord-OCR network (\Cref{sec:impl-chord}) is a second instance
of the same machine-learning pattern applied to a different visual
language, trained on synthetic LilyJAZZ chord strips and fine-tuned
on hand-labelled real strips with a weighted sampler that mixes the
two distributions; the fine-tuning recipe is itself an information-
extraction design pattern (synthetic pretraining followed by
domain-targeted fine-tuning) developed inside the constraints of
the available data.

The full inference pipeline (\Cref{sec:impl-inference}) closes the
loop: it accepts a PDF or image of a previously unseen page,
extracts the symbolic content as structured JSON, and serves the
result through a REST API (\Cref{sec:impl-api}) that demonstrates the
end-to-end automatic extraction of musical information from large
volumes of image data.

% ─────────────────────────────────────────────────────────────────────────────
\section{Methodological Reflection}
\label{sec:conc-reflection}

A short reflection on what worked, what did not, and what would be
done differently closes the assessment.

The synthetic-first methodology paid off.
Re-rendering PrIMuS in the target engraving style cost a one-off
LilyPond setup effort and a few thousand CPU-hours of rendering, but
it gave the project a labelled corpus orders of magnitude larger
than any annotation effort could realistically have produced; this
is the single decision that made the rest of the work feasible
inside a Bachelor's thesis budget.
The pairing of synthetic supervision with a calibrated augmentation
pipeline turned out to be sufficient to bring the scanned-split SER
within \num{0.1} percentage points of the clean-split SER, which
was the most uncertain outcome at the start of the project.

The modular five-stage pipeline (\Cref{sec:design-pipeline}) proved
more valuable in practice than a monolithic end-to-end model would
have been.
Localised failure modes, the ability to debug each stage in
isolation, and the clean replacement contract between stages
(\Cref{tab:stage-contracts}) all simplified iteration.
The deterministic grammar fixer is a particularly favourable
trade-off: a few hundred lines of stateful code remove a class of
errors that would otherwise have required additional model capacity
or constrained decoding to address.

A few decisions would be revisited if the work were restarted.
The CTC mean-logprob rejection threshold was calibrated late in the
project, after a checkpoint had stabilised, and the calibration has
to be redone whenever the recogniser is retrained.
A learned out-of-distribution detector, decoupled from the CRNN's
confidence calibration, would be cleaner.
The chord-OCR network was scoped late as well, after the melody
network had converged; in retrospect the two streams could have been
developed in parallel from the start, which would have given the
chord fine-tuning loop more iteration time.
Finally, the end-to-end measurement on genuine Real Book pages
should have been the first experimental milestone rather than the
last: it would have driven the augmentation calibration earlier and
exposed the visual gaps that the synthetic pipeline does not yet
cover.

The four lessons of the previous version of this chapter survive
intact and are worth recording.
LMX as a model target is decisively more practical than full
MusicXML at the CTC scale.
CTC remains a strong baseline for monophonic OMR despite the
popularity of transformer decoders, particularly inside a thesis
compute budget.
Scan-simulation augmentation in \code{albumentations} closes most of
the synthetic-to-real distribution shift but leaves a measurable
residue on the noisier end of the corpus.
Decoupling melody and chord recognition into two independent CRNNs
prevents vocabulary pollution and admits independent retraining of
either branch, which proved more valuable in practice than the
theoretical appeal of a unified output head.
